% T I T L E   P A G E
% -------------------

\pagestyle{empty}
\pagenumbering{roman}

\begin{titlepage}
        \begin{center}
        \vspace*{0.2cm}

        \Huge
        {\bf Integrating Column-Oriented Storage and Query
        	Processing Techniques into Graph Database
        	Management Systems }

        \vspace*{1.0cm}

        \normalsize
        by \\

        \vspace*{1.0cm}

        \Large
        Pranjal Gupta \\

        \vspace*{3.0cm}

        \normalsize
        A thesis \\
        presented to the University of Waterloo \\ 
        in fulfillment of the \\
        thesis requirement for the degree of \\
        Master of Mathematics \\
        in \\
        Computer Science \\

        \vspace*{2.0cm}

        Waterloo, Ontario, Canada, 2020 \\

        \vspace*{1.0cm}

        \copyright\ Pranjal Gupta 2020 \\
        \end{center}
\end{titlepage}

\pagestyle{plain}
\setcounter{page}{2}

\cleardoublepage

\cleardoublepage

% D E C L A R A T I O N   P A G E
% -------------------------------

\noindent
I hereby declare that I am the sole author of this thesis. This is a true copy of the thesis, including any required final revisions, as accepted by my examiners.

\bigskip
  
\noindent
I understand that my thesis may be made electronically available to the public.

\cleardoublepage

% A B S T R A C T
% ---------------

\begin{center}\textbf{Abstract}\end{center}

Column-oriented RDBMSs, which support traditional read-heavy analytics workloads, employ a set of storage and query processing techniques for scalability and performance, such as positional tuple IDs, column-specific compression, and block-oriented processing. We revisit these techniques in the context of contemporary graph database management systems (GDBMSs). GDMSs support a new set of analytics workloads, such as fraud detection in financial transaction networks or recommendations in social networks, which are also read-heavy but have fundamentally different access patterns than traditional analytics workloads. We first review the properties of data and access patterns of queries in GDBMS to identify the components of GDBMSs where existing columnar techniques can and cannot directly be used. We then present the physical data layout of columnar data structures, new columnar compression, and query-processing techniques that are optimized for GDBMSs. Our techniques include a new compact vertex and edge ID scheme, a new null and empty list compression scheme based on prefix-sums, and list-based query processing. We have integrated our techniques into GraphflowDB, an in-memory GDBMS. Compared to uncompressed storage, our compression techniques has scaled the system by Xx with minimal performance overheads. Our null compression scheme outperforms existing columnar schemes in query performance, with minor loss in compression rate and achieves both higher compression rate and better query performance as compared to row-oriented storage techniques adopted by existing GDBMSs. Finally, our list-based query processor techniques improve query performance by Xx on a variety of path queries and significantly outperform their corresponding conventional versions.

\cleardoublepage

% A C K N O W L E D G E M E N T S
% -------------------------------

\begin{center}\textbf{Acknowledgements}\end{center}

First and foremost, I want to thank my supervisor, Professor Semih Salihoglu, for all the guidance and support he has provided me over the past two years. Working with Semih has been a great experience and am really grateful to be a part of his reaserch group. He has been a constant source of inspiration to whom I can look upto. Thank you for being a great teacher and guide.

I next want to thank Siddhartha, for being a friend to whom I can turn up anytime and for anything, and Amine, for helping me with the Graphflow project, collaborating with me on the A+ Indexes and also coping with my not-so-good reaction at times. All nighters were fun too with you guys. I am also grateful to Manoj, Antony and Aatish who helped me during my time here; and to my friends Yash and Garvit for just being there to listen to me.

I also want to express my sincerest gratitude to my thesis readers, Ken Salem and Tamer Özsu, for taking out valuable time from their busy schedules to read my thesis and provide their valuable comments. 

I would also like to thank my parent for their guidence, support and encouragement in all my endeavours in life; and my siblings, Pranshri and Shrijal, for keeping up my morale. 

Finally, I want to thank my friend, Snehal, for being my constant source of love, happiness and motivation. Thank you for seeing me through all my good and bad phases. 

\cleardoublepage

% D E D I C A T I O N
% -------------------

\begin{center}\textbf{Dedication}\end{center}

This is dedicated to the ones I care about .
\cleardoublepage

% T A B L E   O F   C O N T E N T S
% ---------------------------------
\renewcommand\contentsname{Table of Contents}
\tableofcontents
\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% L I S T   O F   T A B L E S
% ---------------------------
\addcontentsline{toc}{chapter}{List of Tables}
\listoftables
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page

% L I S T   O F   F I G U R E S
% -----------------------------
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page

% GLOSSARIES (Lists of definitions, abbreviations, symbols, etc. provided by the glossaries-extra package)
% -----------------------------
\printglossaries
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page

% Change page numbering back to Arabic numerals
\pagenumbering{arabic}

